%% Write your acknowledgements in this file. If you do not want to acknowledge anyone,
%% you can delete this file and comment out the corresponding part in the "thesis.tex"
%% file.
%

This thesis would not have been possible without the help of many people.
I would like to primarily acknowledge the excellent mentoring of Kyle Cranmer, who has served as my thesis advisor for my graduate school carreer.
In addition, I would like to thank Allen Mincer and Peter Nemethy for for their guidance, advice, and help throughout my time at New York University.
I am indebeted to Akira Shibata for teaching me particle physics, computer programming, and for being an excellent example of a productive and successful scientist.
Much of the analysis done toward my degree (and throughout the ATLAS experiment) depended on his contributions to software and measurement techniques.
I would also like to thank Attila Krasznahorkay for the incrdible amount of help, both technical and conceptual, that he has provided me over the past few years.
Aside from being an excellent mentor, he has developed and maintained much of the software used to process and analyze the incredible amount of data
produced by the ATLAS detector that made this thesis possible.
I would also like acknowledge Jonathan Zrake for always being available for a discussion, either long or short.
And I would also like to thank Dan Foreman-Mackey for helping to open up the world of computer science to me.

Finally, I would like to thank my parents and the rest of my extended family for their continual support.
