
%
\section{Theory}

% Info
% Resonnaances
% http://resonaances.blogspot.com/2011/04/update-on-forward-backward-asymmetry.html

% CDF
% http://arxiv.org/abs/1101.0034

% D0
% 10.1103/PhysRevD.84.112005

% References:
% Top FB Assymetry and Same Sign Production
% http://arxiv.org/abs/1109.3202


% http://xxx.lanl.gov/abs/arXiv:1102.3374


Recent experimental results have found statistically significant deviations from standard model predictions in events involving top quarks.
These deviations suggest that new physical models may play a role in the top sector that is experimentally testable at the LHC.

The Tevatron is a proton anti-proton collider located at Fermilab outide of Chicaco, Illinois.


To lowest order, the pair production of top quarks is symmetric under charge conjugation.
Therefore, at tree level in QCD, one does not expect an angular asymmetry in the pair production of top quarks.
However, at higher order in QCD, the Standard Model predicts a small asymmetry due to loop corrections.
This asymmetry can be measured experimentally using an observable that probes the directions of the top quark pairs produced at the tevatron.
Defining the difference in rapidity between the top and anti-top quarks produced to be $\Delta y = y_{t} - y_{\tbart}$, one can mesure the number of events with $\Delta y>0$ and $\Delta y<0$, which we referr to as $N(\Delta y>0)$ and $N(\Delta y < 0)$, respectively.
Using these definitions, one constructs the total $\tbart$ frame asymmetry as
\begin{equation}
  A^{\tbart} = frac{N(\Delta y>0) - N(\Delta y>0)}{N(\Delta y>0) + N(\Delta y>0)}
\end{equation}

The standard model prediction for $A^{\tbart}$ is on the order of $0.06 \pm 0.01$.%http://arxiv.org/pdf/1101.0034v1.pdf
% [1] L. G. Almeida, G. F. Sterman and W. Vogelsang, Phys. Rev. D 78, 014008 (2008).
% [2] O. Antunano, J. H. Kuhn, and G. V. Rodrigo, Phys. Rev. D 77, 014003 (2008).
% [3] M. T. Bowen, S. D. Ellis, and D. Rainwater, Phys. Rev. D 73, 014008 (2006).

Both D0 and CDF found 


% From HEre: http://arxiv.org/pdf/1101.0034v1.pdf
Selection      A^{\tbart}_{FB}( \Delta Y < 1.0) A^{\tbart}_{FB}( \Delta Y > 1.0)
CDF Result     0.026 \pm 0.118       0.611 \pm 0.256
SM Prediction  0.039 \pm 0.006       0.123 \pm 0.008

Selection      A^{\tbart}_{FB}( M_{\ttbar} < 450 GeV) A^{\tbart}_{FB}( M_{\ttbar} \ge 450 GeV)
CDF Result     -0.116 \pm 0.153      0.475 \pm 0.114
SM Prediction  0.040 \pm 0.006       0.088 \pm 0.013 

CDF : 20.1 \pm 6.7 
D0 : 19.6 \pm 6.5 % D0 http://arxiv.org/abs/1107.4995

blah
